---
title: "Getting started with OtsuFire"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with OtsuFire}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  eval     = requireNamespace("sf", quietly = TRUE) &&
             requireNamespace("terra", quietly = TRUE)
)
```

`OtsuFire` provides a reproducible burned-area mapping pipeline with
four composable layers:

1. A **mosaic** stage that builds an annual change-index raster.
2. A **deterministic** Otsu / seed-and-grow segmentation workflow.
3. A **supervised** one-year probabilistic workflow that learns
   `p_burned` from XGBoost models trained on deterministic decisions.
4. A workflow-independent **validation** utility.

This vignette walks through the API surface using synthetic toy data
that runs in seconds. Operational, full-scale examples live in the
companion vignette `workflow-overview`.

## Public API

```{r api}
library(OtsuFire)

# 16 public exports.
public_api <- c(
  "change_index_mosaic", "validate_fire_maps",
  "build_burned_mapping_config", "detect_burned_patches",
  "score_burned_patches", "run_deterministic_pipeline",
  "build_supervised_burned_config", "build_supervised_training_pools",
  "make_spatial_folds", "extract_supervised_features",
  "run_oof_diagnostics", "train_final_burned_model",
  "score_supervised_burned_map", "update_burned_like_registry",
  "check_supervised_consistency", "run_oneyear_supervised_pipeline"
)
length(public_api)
```

## A toy supervised configuration

The `build_*_config()` builders are pure validators: they normalise
the inputs into S3 objects without reading rasters or training
models. They are the safest entry points for exploring the API.

```{r cfg, eval = requireNamespace("sf", quietly = TRUE) && requireNamespace("terra", quietly = TRUE)}
suppressPackageStartupMessages({
  library(sf)
  library(terra)
})

# Synthetic 8x8 change-index raster + matching burnable mask.
ci  <- tempfile(fileext = ".tif")
terra::writeRaster(
  terra::rast(ncol = 8, nrow = 8, vals = 1:64), ci, overwrite = TRUE
)
mask <- tempfile(fileext = ".tif")
terra::writeRaster(
  terra::rast(ncol = 8, nrow = 8, vals = rep(1, 64)),
  mask, overwrite = TRUE
)

# Synthetic single-polygon "internal_decisions" GPKG.
decisions <- tempfile(fileext = ".gpkg")
sf::st_write(
  sf::st_sf(
    id = 1L,
    geometry = sf::st_sfc(
      sf::st_polygon(list(rbind(c(0,0), c(1,0), c(1,1), c(0,1), c(0,0)))),
      crs = 3035
    )
  ),
  decisions, quiet = TRUE
)

cfg_sup <- build_supervised_burned_config(
  scenario           = "balanced",
  internal_decisions = decisions,
  change_index       = ci,
  target_year        = 2020L,
  output_dir         = tempdir()
)
print(cfg_sup)
```

## Modular building blocks

The supervised workflow is decomposed into modular helpers that each
take a config (or relevant subset) and return either an `sf` object
or a path to an output file. They share the same validator pattern:
calling them with no arguments returns an informative error rather
than a low-level traceback.

```{r modular, error = TRUE}
build_supervised_training_pools()
make_spatial_folds()
extract_supervised_features()
run_oof_diagnostics()
train_final_burned_model()
score_supervised_burned_map()
update_burned_like_registry()
check_supervised_consistency()
```

These messages document what each function expects without
triggering any heavy computation.

## Where to go next

* `vignette("workflow-overview", package = "OtsuFire")` — explains
  how the four layers compose into a full one-year run, with
  references to the relevant config options and outputs.
* The function reference (run `?OtsuFire::run_oneyear_supervised_pipeline`)
  documents every exported function with parameter-level detail.
* The `HANDOFF_SUMMARY_FOR_NEW_AGENT.md` file in the repository
  documents the design decisions and validated baselines behind the
  current scientific defaults.
