---
title: "Getting started with OtsuFire"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with OtsuFire}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  eval     = requireNamespace("sf", quietly = TRUE) &&
             requireNamespace("terra", quietly = TRUE)
)
```

`OtsuFire` is an open-source R package for long-term, wall-to-wall
burned-area reconstruction. It implements a self-labelling, weakly
supervised framework that generates training labels without requiring
prior knowledge of fire locations or reliable target-year reference data.

The framework separates **supervision generation** from **final
classification**: Otsu-guided segmentation and physically interpretable
rules provide the evidence used to construct training pools, which
then train a probabilistic patch-level classifier.

The package provides four composable stages:

1. **Mosaic preparation:** assemble image tiles into an annual
   change-index raster.

2. **Otsu-guided segmentation and rule-based filtering:** delineate
   candidate patches using Otsu thresholds and seed-and-grow
   segmentation, then apply spectral, spatial and contextual filters
   to assign `keep`, `review` or `drop` decisions.

3. **Probabilistic refinement:** construct burned and unburned training
   pools, train an XGBoost classifier, and estimate `p_burned` for
   candidate patches to produce the final burned-area map.

4. **Independent validation:** compare mapped burned areas with an
   external reference. This utility can also evaluate maps generated
   outside OtsuFire.

Wall-to-wall coverage describes the mapping domain: the whole study area
is searched for candidate patches. Probabilistic scoring then operates on
those candidate patches.

## Package, workflow book and data

The OtsuFire package and its source code are available on
[GitHub](https://github.com/olgaviedma/OtsuFire).

For step-by-step guidance on implementing and using OtsuFire, see the
[OtsuFire workflow book](https://github.com/nataliaquintero/OtsuFire-book).
The companion repository also includes the Google Earth Engine scripts
used for image selection, masking and compositing.

Burned-area maps for the full time series are available on
[Zenodo](https://doi.org/10.5281/zenodo.22893763).

## Understanding the function names

Before exploring the code, it helps to clarify two terms used in the API.

**`deterministic`** refers to Otsu-guided segmentation and rule-based
filtering, as in `run_deterministic_pipeline()`. This stage follows
explicit algorithms and decision rules without fitting a classifier
to labelled examples. The thresholds are not necessarily fixed:
Otsu thresholds are estimated from the input data within the defined
environmental strata.

**`supervised`** refers to the classifier trained on burned and unburned
examples, as in `run_oneyear_supervised_pipeline()`. Its labels are
generated internally from the preceding evidence. This is why the
overall framework is described as **weakly supervised**: training
relies on labels inferred from rules and indirect evidence, which
may contain errors, rather than independently verified labels for
every example.

The training pools distinguish high-confidence burned patches,
unburned background, and moderately and strongly burned-like
unburned surfaces. These last two pools help the classifier learn
to distinguish burning from confounding changes. All three unburned
pools share the same unburned class label.

`review` patches are withheld from training but remain available for
prediction. The classifier scores the complete candidate-patch set
using descriptors extracted from the underlying data, excluding
rule-based decisions and their associated support metrics from its
predictors. The final classification therefore goes beyond simply
retaining the initial `keep` patches: `keep`, `review` and `drop` are
rule-based decisions, not the classes of the final map.

## What this vignette covers

Let's start with the public API and a small configuration example
using synthetic data. These examples introduce the package's building
blocks without running a full mapping analysis or training a model.

For a complete one-year workflow, including training-pool construction
and validation, see the companion vignette `workflow-overview`.

## Public API

```{r api}
library(OtsuFire)

public_api <- sort(getNamespaceExports("OtsuFire"))
length(public_api)
public_api
```

The exports fall into five groups. The chunk below lists them and checks
that the groups cover exactly the current exports, so this list cannot
drift from the package:

```{r api-groups}
api_groups <- list(
  "Mosaic preparation" = c(
    "change_index_mosaic", "mosaic_from_tiles", "clean_raster_file"
  ),
  "Otsu-guided segmentation and rule-based filtering (deterministic API)" = c(
    "build_burned_mapping_config", "run_deterministic_pipeline",
    "detect_burned_patches", "score_burned_patches",
    "apply_chain_cleaning", "apply_chain",
    "collect_keep_reference_samples", "build_keep_pool_from_samples"
  ),
  "Probabilistic refinement (supervised API)" = c(
    "build_supervised_burned_config", "validate_supervised_execution",
    "build_supervised_training_pools", "make_spatial_folds",
    "extract_supervised_features", "run_oof_diagnostics",
    "train_final_burned_model", "score_supervised_burned_map",
    "write_thresholded_burned", "run_oneyear_supervised_pipeline"
  ),
  "Training-pool curation" = c(
    "diagnose_training_pools", "export_visual_validation_pools",
    "apply_visual_validation", "promote_artifact_hard_negatives",
    "subsample_random_background", "add_pool_shape_features",
    "assemble_era_training_pool"
  ),
  "Validation and consistency" = c(
    "validate_fire_maps", "check_supervised_consistency"
  )
)
stopifnot(setequal(unlist(api_groups), public_api))
str(api_groups)
```

`apply_chain()` is an alias of `apply_chain_cleaning()`: both names refer
to the same function.

```{r alias}
identical(apply_chain, apply_chain_cleaning)
```

The two functions in the last group answer different questions.
`validate_fire_maps()` assesses accuracy against an **external** reference
of burned areas. `check_supervised_consistency()` compares the outputs of
the two mapping stages of the same run; agreement between pipeline stages
is not independent validation.

## Build your first supervised-stage configuration

We'll start by building a configuration: an object that stores the
inputs and settings for the supervised stage. Creating this object
does not train a classifier or produce a burned-area map.

`build_supervised_burned_config()` checks the arguments (types, allowed
values and the target year), confirms that the optional input files it is
given exist, and resolves the default settings and output locations. It
does not read the contents of the rasters or vectors, and it does not
create any folders; those checks happen in the workflow stages and in
`validate_supervised_execution()`.

The synthetic inputs below illustrate the configuration only. Their
values do not represent realistic fire observations. They are written to
a temporary folder:

* an 8 x 8 change-index raster in a projected CRS (ETRS89-LAEA,
  EPSG:3035) with an explicit extent;
* a burnable mask derived from that raster, so both share the same grid;
* a one-polygon decision layer, in the same CRS and overlapping the
  raster, stored in a GeoPackage layer named `internal_decisions`.

The decision layer carries a `class_final` column, the label field the
supervised workflow expects. It is enough for building a configuration,
but it is not a valid training dataset: the full workflow needs a real
decision layer produced by the Otsu-guided segmentation stage.

```{r cfg}
work_dir <- tempfile("otsufire_vignette_")
dir.create(work_dir)

# Change-index raster: 8 x 8 cells of 100 m in EPSG:3035.
ci <- terra::rast(
  nrows = 8, ncols = 8,
  xmin = 3000000, xmax = 3000800,
  ymin = 2000000, ymax = 2000800,
  crs = "EPSG:3035"
)
terra::values(ci) <- seq(0, 630, by = 10)
ci_path <- file.path(work_dir, "change_index.tif")
terra::writeRaster(ci, ci_path, overwrite = TRUE)

# Burnable mask on the same grid (1 = burnable).
burnable <- terra::ifel(is.na(ci), 0, 1)
burnable_path <- file.path(work_dir, "burnable_mask.tif")
terra::writeRaster(burnable, burnable_path, overwrite = TRUE)

# One candidate patch inside the raster extent, same CRS.
decisions <- sf::st_sf(
  poly_id = 1L,
  class_final = "keep",
  geometry = sf::st_sfc(
    sf::st_polygon(list(rbind(
      c(3000200, 2000200), c(3000600, 2000200),
      c(3000600, 2000600), c(3000200, 2000600),
      c(3000200, 2000200)
    ))),
    crs = 3035
  )
)
decisions_path <- file.path(work_dir, "internal_decisions.gpkg")
sf::st_write(decisions, decisions_path,
             layer = "internal_decisions", quiet = TRUE)

cfg_sup <- build_supervised_burned_config(
  run_label          = "balanced",
  internal_decisions = decisions_path,
  change_index       = ci_path,
  burnable_mask      = burnable_path,
  target_year        = 2020L,
  output_dir         = file.path(work_dir, "results")
)
```

The printed configuration shows the resolved settings and inputs. To keep
the output readable, the temporary-folder prefix of the paths is displayed
as `<tempdir>`; this substitution affects the display only, not the paths
stored in `cfg_sup`.

```{r cfg-print}
printed <- capture.output(print(cfg_sup))
tmp_prefixes <- unique(c(
  normalizePath(tempdir(), winslash = "/", mustWork = FALSE),
  gsub("\\\\", "/", tempdir())
))
for (p in tmp_prefixes) printed <- gsub(p, "<tempdir>", printed, fixed = TRUE)
cat(printed, sep = "\n")
```

## Explore the individual workflow steps

The supervised workflow can also be run step by step: construct training
pools, create spatial folds, extract predictors, inspect out-of-fold
diagnostics, fit the final model, and score candidate patches. The
function reference describes the inputs and outputs of each step.

The calls below intentionally omit required arguments to illustrate the
input checks. They do not execute the workflow or demonstrate model
performance. Each error is caught locally and its message is printed.

```{r modular}
show_input_check <- function(expr) {
  tryCatch(
    {
      force(expr)
      message("No error raised.")
    },
    error = function(e) cat("Error:", conditionMessage(e), "\n")
  )
}

show_input_check(build_supervised_training_pools())
show_input_check(make_spatial_folds())
show_input_check(extract_supervised_features())
show_input_check(run_oof_diagnostics())
show_input_check(train_final_burned_model())
show_input_check(score_supervised_burned_map())
show_input_check(check_supervised_consistency())
```

```{r cleanup, include = FALSE}
unlink(work_dir, recursive = TRUE)
```

## Where to go next

* `vignette("workflow-overview", package = "OtsuFire")` walks through a
  complete one-year run: training-pool construction, the supervised stage
  and validation.
* The function reference documents every exported function with
  parameter-level detail; start with `?OtsuFire` or, for example,
  `?run_oneyear_supervised_pipeline`.
* `NEWS.md`, also available with `news(package = "OtsuFire")`, summarises
  what changed in each release.

When you evaluate results, keep two kinds of assessment apart:

* **Internal out-of-fold diagnostics** evaluate predictions on held-out,
  internally generated labels. They measure how well the classifier
  reproduces those labels; they do not verify the true burned status of
  the training labels.
* **External validation** with `validate_fire_maps()` compares the
  resulting map against an independent burned-area reference, within the
  area where that reference provides valid coverage.
