---
title: "Getting started with OtsuFire"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with OtsuFire}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  eval     = requireNamespace("sf", quietly = TRUE) &&
             requireNamespace("terra", quietly = TRUE)
)
```

`OtsuFire` provides a reproducible burned-area mapping pipeline with
four composable layers:

1. A **mosaic** stage that builds an annual change-index raster.
2. A **deterministic** Otsu / seed-and-grow segmentation workflow.
3. A **supervised** one-year probabilistic workflow that learns
   `p_burned` from XGBoost models trained on deterministic decisions.
4. A workflow-independent **validation** utility.

This vignette walks through the API surface using synthetic toy data
that runs in seconds. Operational, full-scale examples live in the
companion vignette `workflow-overview`.

## Public API

```{r api}
library(OtsuFire)

# Public exports (see the package NAMESPACE for the authoritative list).
public_api <- sort(getNamespaceExports("OtsuFire"))
length(public_api)
public_api
```

The exports cover:

* the mosaic stage: `change_index_mosaic`, `mosaic_from_tiles`,
  `clean_raster_file`;
* the deterministic stage: `build_burned_mapping_config`,
  `detect_burned_patches`, `score_burned_patches`,
  `run_deterministic_pipeline`, `apply_chain_cleaning` (alias
  `apply_chain`), `collect_keep_reference_samples`,
  `build_keep_pool_from_samples`;
* the supervised stage: `build_supervised_burned_config`,
  `validate_supervised_execution`, `build_supervised_training_pools`,
  `make_spatial_folds`, `extract_supervised_features`,
  `run_oof_diagnostics`, `train_final_burned_model`,
  `score_supervised_burned_map`, `write_thresholded_burned`,
  `run_oneyear_supervised_pipeline`;
* training-pool curation: `diagnose_training_pools`,
  `export_visual_validation_pools`, `apply_visual_validation`,
  `promote_artifact_hard_negatives`, `subsample_random_background`,
  `add_pool_shape_features`, `assemble_era_training_pool`;
* validation and consistency: `validate_fire_maps`,
  `check_supervised_consistency`.

## A toy supervised configuration

The `build_*_config()` builders are pure validators: they normalise
the inputs into S3 objects without reading rasters or training
models. They are the safest entry points for exploring the API.

```{r cfg, eval = requireNamespace("sf", quietly = TRUE) && requireNamespace("terra", quietly = TRUE)}
suppressPackageStartupMessages({
  library(sf)
  library(terra)
})

# Synthetic 8x8 change-index raster + matching burnable mask.
ci  <- tempfile(fileext = ".tif")
terra::writeRaster(
  terra::rast(ncol = 8, nrow = 8, vals = 1:64), ci, overwrite = TRUE
)
mask <- tempfile(fileext = ".tif")
terra::writeRaster(
  terra::rast(ncol = 8, nrow = 8, vals = rep(1, 64)),
  mask, overwrite = TRUE
)

# Synthetic single-polygon "internal_decisions" GPKG.
decisions <- tempfile(fileext = ".gpkg")
sf::st_write(
  sf::st_sf(
    id = 1L,
    geometry = sf::st_sfc(
      sf::st_polygon(list(rbind(c(0,0), c(1,0), c(1,1), c(0,1), c(0,0)))),
      crs = 3035
    )
  ),
  decisions, quiet = TRUE
)

cfg_sup <- build_supervised_burned_config(
  run_label          = "balanced",
  internal_decisions = decisions,
  change_index       = ci,
  target_year        = 2020L,
  output_dir         = tempdir()
)
print(cfg_sup)
```

## Modular building blocks

The supervised workflow is decomposed into modular helpers that each
take a config (or relevant subset) and return either an `sf` object
or a path to an output file. They share the same validator pattern:
calling them with no arguments returns an informative error rather
than a low-level traceback.

```{r modular, error = TRUE}
build_supervised_training_pools()
make_spatial_folds()
extract_supervised_features()
run_oof_diagnostics()
train_final_burned_model()
score_supervised_burned_map()
check_supervised_consistency()
```

These messages document what each function expects without
triggering any heavy computation.

## Where to go next

* `vignette("workflow-overview", package = "OtsuFire")` — explains
  how the four stages combine into a full one-year run, the training
  pools, and the difference between the internal (OOF) and external
  (`validate_fire_maps()`) validations.
* The function reference (for example
  `?OtsuFire::run_oneyear_supervised_pipeline`) documents every exported
  function with parameter-level detail.
* `NEWS.md` summarises what changed in each release.
