---
title: "OtsuFire workflow overview"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{OtsuFire workflow overview}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
# eval = FALSE silences knitting: the examples need full-scale data.
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>", eval = FALSE
)
```

## Package, workflow book and data

The OtsuFire package and its source code are available on
[GitHub](https://github.com/olgaviedma/OtsuFire).

For step-by-step guidance on implementing and using OtsuFire, see the
[OtsuFire workflow book](https://github.com/nataliaquintero/OtsuFire-book).
The companion repository also includes the Google Earth Engine scripts
used for image selection, masking and compositing.

Burned-area maps for the full time series are available on
[Zenodo](https://doi.org/10.5281/zenodo.22893763).

Tables S13–S15 of the accompanying manuscript's supplementary material
describe the functions included in the package.

## Introduction

OtsuFire combines four stages: **mosaic preparation**, **Otsu-guided
segmentation and rule-based filtering**, **probabilistic refinement**,
and **independent validation**.

The workflow separates training-label generation from final
classification. Segmentation and filtering provide the evidence used
to construct training pools; an XGBoost classifier then learns from
these pools to estimate a burned score for each candidate patch.
This internally generated supervision makes OtsuFire a
**weakly supervised** framework.

In the API, `deterministic` identifies the segmentation and filtering
stage, which follows explicit algorithms and rules, including
data-adaptive Otsu thresholds. `supervised` identifies the classifier
trained on the resulting labels.

This vignette summarises how the stages connect; the workflow book above
covers data preparation, configuration and interpretation of results
step by step.

The examples below are **not evaluated** and require your own data
and paths. For a small synthetic configuration example, see
`vignette("OtsuFire", package = "OtsuFire")`.

## 1. Prepare the mosaic

`change_index_mosaic()` combines change-index tiles, such as RBR or
dNBR, into a single annual raster. This becomes the `change_index`
input for the following stages.

```{r mosaic, eval=FALSE}
library(OtsuFire)

mosaic_res <- change_index_mosaic(
  folder_path    = "/path/to/yearly/tiles",
  mask           = "/path/to/study_area.gpkg",
  year           = 2020L,
  raster_pattern = "RBR_2020_*.tif",
  output_dir     = "/path/to/mosaics",
  nodata_value   = -9999,
  cap_below      = TRUE,
  lower_cap      = -1000,
  method         = "merge"
)

mosaic_path <- mosaic_res$mosaic_path
```

Use a folder containing the intended input tiles, and set
`raster_pattern` to match their file names. See `?change_index_mosaic`
for file-selection and mosaicking options.

## 2. Generate and filter candidate patches

Otsu-guided segmentation identifies seed pixels and grows candidate
patches. Rule-based filters then assess spectral, spatial and contextual
evidence, assigning each candidate to `keep`, `review` or `drop`.

The main output is `internal_decisions.gpkg`, which links candidate
geometry to the filtering decisions used in training-pool construction.

```{r segmentation, eval=FALSE}
det_cfg <- build_burned_mapping_config(
  change_index   = mosaic_path,
  vegetation_map = "/path/to/vegetation_classes.tif",
  burnable_mask  = "/path/to/burnable_mask.tif",
  target_year    = 2020L,
  output_dir     = "/path/to/results",
  run_name       = "rbr_2020"
)

det_run <- run_deterministic_pipeline(det_cfg)

decisions_path <- det_run$result_paths$internal_decisions
```

This example shows the main connection between stages. Configure
additional spatial inputs and detection, refinement and scoring
settings for your study area using `?build_burned_mapping_config`
and the book.

## 3. Refine the classification

The supervised stage constructs training pools, extracts predictors,
runs out-of-fold diagnostics, and fits an XGBoost classifier. It then
scores the complete candidate-patch set.

```{r supervised, eval=FALSE}
sup_cfg <- build_supervised_burned_config(
  run_label          = "balanced",
  internal_decisions = decisions_path,
  change_index       = mosaic_path,
  hotspots           = "/path/to/hotspots_2020.gpkg",
  target_year        = 2020L,
  output_dir         = "/path/to/results",
  options = list(
    data_base      = "/path/to/data",
    composite_base = "/path/to/composites"
  )
)

sup_run <- run_oneyear_supervised_pipeline(
  config          = sup_cfg,
  run_consistency = TRUE,
  overwrite       = FALSE
)
```

The configuration above illustrates the API connection; it does not
specify every setting used in the manuscript. See the book for the
complete configuration and `?build_supervised_burned_config` for
supported inputs and defaults.

### Training pools at a glance

The burned pool comes from high-confidence `keep` patches. Depending
on the configuration, unburned examples come from three sources:

| Pool | Internal name | Purpose |
|------|---------------|---------|
| Burnable background | `random` | Represent low-ambiguity unburned conditions. |
| Moderately burned-like | `otsu` | Represent rejected candidates that resemble burns. |
| Strongly burned-like | `artifact_hard` | Represent more challenging burned-like confounders. |

All three negative pools receive the same unburned class label.
Pool selection, sampling and balancing are configurable.

`review` patches are excluded from training but remain available for
prediction. The final classification can therefore differ from the
initial rule-based decisions.

Burned scores should not be interpreted as independently calibrated
probabilities. The selected score field and classification threshold
must be consistent with the intended final product.

## 4. Assess the results

OtsuFire provides three complementary checks:

| Assessment | What it measures |
|------------|------------------|
| Internal OOF diagnostics | Performance on held-out, internally generated labels. |
| External validation | Agreement with an independent burned-area reference. |
| Consistency checks | Differences between the segmentation/filtering and supervised outputs. |

OOF diagnostics assess learning from the generated labels; they do not
independently establish whether those labels are correct. Consistency
between stages also does not establish mapping accuracy.

### External validation

`validate_fire_maps()` compares an already-thresholded burned-area map
with an external reference within the validation domain. It reports
metrics including omission, commission, F1 and IoU.

Prepare the binary burned-area layer using the score and threshold
selected for your run before calling this function.

```{r validation, eval=FALSE}
val <- validate_fire_maps(
  input_shapefile = "/path/to/thresholded_burned_2020.gpkg",
  ref_shapefile   = "/path/to/reference_fires_2020.gpkg",
  mask_shapefile  = "/path/to/validation_mask_2020.gpkg",
  burnable_raster = "/path/to/burnable_mask.tif",
  year_target     = 2020L,
  validation_dir  = "/path/to/results/validation",
  metrics_type    = "all",
  observability_raster = "/path/to/observation_doy_2020.tif",
  observability_mode   = "wholefire_fraction",
  ref_obs_doy_col      = "obs_required_doy"
)

val$metrics
```

**Temporal observability** checks whether the imagery could capture
each reference fire. The example uses `wholefire_fraction`, which
assesses the fraction of each fire with suitable observation dates.
Observability is handled at the reference-fire level rather than
by trimming individual pixels within retained fires.

Choose the date fields, required observable fraction and missing-date
policy to match your reference data. See `?validate_fire_maps` and the
book for these settings and the associated audit outputs.

### Consistency between stages

`check_supervised_consistency()` compares the rule-based decisions
with the supervised outputs from the same run.

```{r consistency, eval=FALSE}
cons <- check_supervised_consistency(
  deterministic_decisions = decisions_path,
  final_map               = sup_run$final_map_gpkg,
  config                  = sup_cfg
)
```

This comparison helps inspect how probabilistic refinement changes
the initial decisions. It is separate from external validation.

## Main outputs

The workflow produces:

- Candidate patches and their `keep`, `review` or `drop` decisions.
- Burned and unburned training pools.
- Spatial folds, extracted predictors and OOF diagnostics.
- A fitted classifier and scored candidate patches.
- A final burned-area map after thresholding.
- Validation metrics and diagnostic files when requested.

The `DETERMINISTIC/` and `SUPERVISED/` output folders retain the API
terminology explained above. Use the paths returned by the pipeline
functions to locate the products of each run.

## Continue with the step-by-step tutorial

The [OtsuFire book](https://github.com/nataliaquintero/OtsuFire-book)
provides the step-by-step workflow, from preparing inputs and selecting
settings to inspecting and validating the final maps.

For individual arguments and return values, consult:

```{r help, eval=FALSE}
?change_index_mosaic
?build_burned_mapping_config
?build_supervised_burned_config
?validate_fire_maps
```
