---
title: "OtsuFire workflow overview"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{OtsuFire workflow overview}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
# eval = FALSE silences knitting. purl = FALSE is set per-chunk below
# instead of globally — opts_chunk$set(purl=FALSE) is not honoured by
# knitr::purl() because the setup chunk is extracted as text, not
# evaluated, during the R CMD check "running R code from vignettes"
# stage.
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>", eval = FALSE
)
```

This vignette describes the four stages of the OtsuFire workflow at the
API level: the mosaic stage, the **Otsu-guided segmentation** stage, the
**probabilistic refinement** stage and validation. In the code, the
Otsu-guided segmentation stage is run with `run_deterministic_pipeline()`
and writes to `DETERMINISTIC/`; the probabilistic refinement stage, which
trains a supervised XGBoost model, is run with
`run_oneyear_supervised_pipeline()` and writes to `SUPERVISED/`.

Code chunks are shown but **not evaluated**, because the Otsu-guided
segmentation and probabilistic refinement workflows operate on full-scale
annual mosaics that are too large to ship with a CRAN package. The toy
examples in `vignette("OtsuFire", package = "OtsuFire")` cover the API
without external data.

## 1. Mosaic stage

`change_index_mosaic()` builds an annual change-index raster (for example
RBR or dNBR) from raster tiles. The output is a single GeoTIFF per year,
which becomes the `change_index` input of the Otsu-guided segmentation and probabilistic refinement
configurations.

```{r mosaic, purl = FALSE}
library(OtsuFire)

mosaic_res <- change_index_mosaic(
  folder_path    = "/path/to/yearly/tiles",
  mask           = "/path/to/study_area.gpkg",
  year           = 2020L,
  raster_pattern = "RBR_2020_*.tif",
  output_dir     = "/path/to/mosaics",
  nodata_value   = -9999,
  cap_below      = TRUE,
  lower_cap      = -1000,
  method         = "merge"
)

mosaic_path <- mosaic_res$mosaic_path
```

## 2. Otsu-guided segmentation stage

The Otsu-guided segmentation workflow estimates Otsu thresholds per land-cover and
ecoregion unit, grows seed-supported candidate patches, applies rule-based
filters, and writes a decision layer (`internal_decisions.gpkg`) in which
every candidate patch is classified as `keep`, `review` or `drop`.

```{r det, purl = FALSE}
det_cfg <- build_burned_mapping_config(
  change_index   = mosaic_path,
  vegetation_map = "/path/to/vegetation_classes.tif",
  burnable_mask  = "/path/to/burnable_mask.tif",
  target_year    = 2020L,
  output_dir     = "/path/to/results",
  run_name       = "rbr_2020_custom",
  detect_params = list(
    seed_threshold = 310,
    growth_delta = 90,
    minimum_growth_threshold = 240,
    minimum_seed_pixels = 30L
  ),
  refine_params = list(
    aoi_buffer_m = 5000,
    minimum_detected_area_m2 = 10,
    merge_overlaps = TRUE
  ),
  scoring_params = list(
    support_buffer_m = 0,
    previous_fire_exclusion_buffer_m = 90,
    previous_fire_cleanup_buffer_m = 90,
    minimum_remaining_area_m2 = 20000,
    reference_buffer_m = 90
  )
)

det_run <- run_deterministic_pipeline(det_cfg)
det_run$result_paths$internal_decisions
```

The detection, refinement and scoring settings are passed through
`detect_params`, `refine_params` and `scoring_params`. Use the
`*_by_vegetation` keys when seed or growth behaviour must differ by
land-cover class; see `?build_burned_mapping_config` for the
vegetation-class table and the meaning of each parameter.

For step-by-step control, the Otsu-guided segmentation stage can also be run as two
separate steps:

```{r det-modular, purl = FALSE}
patches <- detect_burned_patches(det_cfg)
scored  <- score_burned_patches(patches$refined_patches_path, det_cfg)
# scored$internal_decisions is an sf with class_final, filter_1..3, etc.
```

## 3. One-year probabilistic refinement stage

The probabilistic refinement stage trains an XGBoost model on labels derived from the
Otsu-guided segmentation decisions, runs out-of-fold (OOF) diagnostics, and writes a
`final_map.gpkg` whose `p_burned` field is the per-patch burned **model
score**.

```{r sup, purl = FALSE}
sup_cfg <- build_supervised_burned_config(
  run_label          = "balanced",
  internal_decisions = det_run$result_paths$internal_decisions,
  change_index       = mosaic_path,
  hotspots           = "/path/to/hotspots_2020.gpkg",
  target_year        = 2020L,
  output_dir         = "/path/to/results",
  options = list(
    data_base      = "/path/to/data",
    composite_base = "/path/to/composites"
  )
)

sup_run <- run_oneyear_supervised_pipeline(
  config           = sup_cfg,
  run_consistency  = TRUE,
  overwrite        = FALSE
)
```

**Training pools.** The burned training pool comprises the deterministic
keep patches and is hereafter referred to as *burned*. Unburned labels
come from up to three differentiated pools that distinguish generic
burnable-background conditions from moderately and strongly burned-like
confounding surfaces (internal pool names in brackets):

* **Burnable background** (`random`) — low-ambiguity unburned cells drawn
  from the burnable domain with a weak immediate change-index response
  (values at or below the 50th percentile of their distribution within the
  sampling domain) and located outside a 500-m exclusion zone around all
  Otsu-guided candidate patches (keep, review and drop).
* **Moderately burned-like** (`otsu`) — candidate patches segmented by Otsu
  as potentially burned but assigned to the drop class by the rule-based
  filters, with `S_PATCH_PA <= 0.15`: the lower end of the burned-like
  confounder spectrum.
* **Strongly burned-like** (`artifact_hard`) — drop patches with a stronger
  burned-like response (high `rbr_med`, low persistence), and therefore
  greater potential to generate commission errors. This pool is **off by
  default** (`negative_pool_params$artifact_hard$enabled = FALSE`). When a
  visual validation is applied with `apply_visual_validation()`, only the
  visually confirmed patches (`VISUAL = 1`) enter training.

Review patches are withheld from training and remain in the complete
candidate-patch universe for subsequent scoring. To control class
imbalance, the burnable-background and moderately burned-like pools are
capped relative to the size of the burned pool
(`negative_pool_params$caps = c(random = 1.0, otsu = 1.0)` by default; an
`Inf` cap disables it), whereas all eligible strongly burned-like patches
are retained and balanced by weight
(`negative_pool_params$artifact_hard$total_weight_ratio`) rather than
capped. The background sample is drawn with the seed `random_seed`, and a
location that is both a background cell and an Otsu patch is kept only
once, as an Otsu patch.

Training uses explicit labels only: a patch becomes a negative only when it
belongs to one of the unburned pools above. Review patches and rows with
missing or unknown labels never become negatives by default. The OOF
diagnostics and the final model use the same eligibility rules, the same
caps and the same feature recipe, so the diagnostics describe the model
that is finally fitted.

Both the OOF models and the final model are trained the same way: the
number of boosting rounds is selected with an inner validation split and
early stopping, and the model is then refitted on all available training
data at that number of rounds.

`p_burned` (and the temporally adjusted `p_burned_current_year`) is a
model score, not a calibrated probability.

Each probabilistic refinement stage can also be run on its own:
`build_supervised_training_pools()`, `make_spatial_folds()`,
`extract_supervised_features()`, `run_oof_diagnostics()`,
`train_final_burned_model()` and `score_supervised_burned_map()`. See
`?build_supervised_burned_config` for all settings.

## 4. Validation and consistency

There are two distinct validations, against two different references.

**Internal validation (OOF).** The out-of-fold diagnostics (`05_OOF`)
measure how well the model reproduces the labels derived from the
Otsu-guided segmentation decisions, using spatial block folds. Each held-out fold is
predicted by a model that never saw it: the outer test fold is not used for
imputation, feature selection, early stopping or fitting. OOF AUC / PR-AUC
are **not ground truth**: the labels are derived from the deterministic
decisions, not from field observations.

**External validation.** `validate_fire_maps()` is independent of the
workflow: threshold `p_burned` (or `p_burned_current_year`) into a
burned-area layer and compare it with an external reference, such as
EFFIS. It reports omission, commission, F1 and IoU.

`validate_fire_maps()` works on an already-thresholded layer, so the scored
map is cut first. The operating threshold is a per-year decision; `0.5`
below is only a placeholder.

```{r validate, purl = FALSE}
fm <- sf::st_read(sup_run$final_map_gpkg, layer = "final_map_full", quiet = TRUE)
burned <- fm[is.finite(fm$p_burned_model) & fm$p_burned_model >= 0.5, ]

thresholded_gpkg <- "/path/to/results/thresholded_burned.gpkg"
sf::st_write(burned, thresholded_gpkg, layer = "thresholded_burned", quiet = TRUE)

val <- validate_fire_maps(
  input_shapefile = thresholded_gpkg,
  ref_shapefile   = "/path/to/reference_fires_2020.gpkg",
  mask_shapefile  = "/path/to/validation_mask_2020.gpkg",
  burnable_raster = "/path/to/burnable_mask.tif",
  year_target     = 2020L,
  validation_dir  = "/path/to/results/validation",
  metrics_type    = "all",
  observability_raster = "/path/to/observation_doy_2020.tif",
  observability_mode   = "wholefire_fraction",
  ref_obs_doy_col      = "obs_required_doy"
)

val$metrics
```

External metrics are computed against a *different reference* than the
OOF metrics. They are **not comparable** with OOF AUC / PR-AUC; keep the
two evaluations distinct.

**Temporal observability.** When an `observability_raster` (a day-of-year
raster) is supplied, observability is assessed **per reference fire**: a
fire is either evaluated in full or excluded in full, and it is never
trimmed at pixel level. Two rules are available through
`observability_mode`.

`"legacy_any"` (the default) excludes a fire whose date falls after the
last observation inside it, or that has no valid observable DOY. Because
the rule compares the fire date with the *maximum* DOY in the perimeter, a
single observable pixel is enough: a fire observed on 1 % of its area is
kept whole. Excluded fires are removed from the reference before
rasterisation, so they contribute no TP or FN, but their territory stays in
the evaluation domain as unburned reference; a correct detection over such
a fire is therefore scored as commission.

`"wholefire_fraction"` requires `temporal_observable_fraction >=
observability_min_fraction` (default `0.75`): the share of the fire's
burnable-domain pixels whose composite DOY is at or after the fire's
required DOY. The required DOY is taken from `ref_obs_doy_col` when that
argument is supplied and from `ref_end_doy_col` otherwise, with **no**
fallback to the start DOY; a fire with no usable date is reported as
`UNDETERMINED_OBS_DATE`.

`ref_obs_doy_col` is **authoritative**: when supplied it is the only date
column consulted, so an `NA` there means "undetermined" and is not
replaced by `end_doy`. `ref_obs_doy_fallback = "end_doy"` allows that
fallback, for reference layers whose observability column is filled for
some sources and not others. `undetermined_cause` records why each fire is
undetermined, and a warning reports how many fires are affected.

A missing date is not evidence that a fire was not observable. With the
default `observability_undetermined_policy = "evaluate"`, fires with
`UNDETERMINED_OBS_DATE` stay in the evaluation as positive reference fires
and are flagged for audit; no date is imputed. `"exclude"` removes them
like a non-observable fire, for sensitivity analyses.

Fires that are not observable (`NOT_OBSERVABLE`, below the threshold, and
`NO_OBSERVABILITY_DATA`, with no observation inside) are removed from the
evaluation domain entirely, so they produce neither omission nor
commission. `OUTSIDE_BURNABLE_DOMAIN` marks fires with no burnable-domain
cell at all. Territory outside these fires is evaluated normally.
`observability_summary` and `OBSERVABILITY_SUMMARY_<year>.csv` report the
counts, the areas by status and the resulting `Evaluable_Area_ha`.

Two CSVs in `02_OBSERVABILITY/` give the per-fire audit trail:
`REFERENCE_FIRES_DETECTION_<year>_<input>.csv` and, for undetermined fires
only, `UNDETERMINED_FIRES_DETECTION_<year>_<input>.csv`. Each row puts the
observability verdict next to what the prediction did on that fire
(reference, detected and omitted cells and area, TP and FN, fire-level
recall), so an individual fire can be revisited without recomputing the
validation.

Partial observability inside kept fires is reported for audit
(`observability_audit`, `OBSERVABILITY_*` CSVs) but never changes the
metrics. Cached reference products are rebuilt automatically when a
relevant input or setting changes; use `force_reprocess_ref = TRUE` to
force a rebuild.

`check_supervised_consistency()` compares the Otsu-guided segmentation and probabilistic refinement
outputs of the same run and writes diagnostic GPKG, CSV and TXT files:

```{r consistency, purl = FALSE}
cons <- check_supervised_consistency(
  deterministic_decisions = det_run$result_paths$internal_decisions,
  final_map               = sup_run$final_map_gpkg,
  config                  = sup_cfg
)
```

The same check runs automatically when
`run_oneyear_supervised_pipeline(..., run_consistency = TRUE)` is called
and `options$data_base` is supplied in the configuration.

## Outputs

A one-year probabilistic refinement run writes, under
`<output_dir>/<target_year>/<run_name>/SUPERVISED/<run_label>/`:

* `01_POOLS/...gpkg` — burned and unburned training pools.
* `02_FOLDS/...gpkg` — spatial folds for OOF.
* `03_FEATURES/features_geometry.gpkg` — full feature table.
* `04_MATRIX/...rds` — XGBoost design-matrix snapshot.
* `05_OOF/..._oof_agg.csv` — out-of-fold diagnostics.
* `07_FINAL_MODEL_V2/..._final_model.rds` — trained model.
* `09_FINAL_MAP/..._final_map.gpkg` — final per-patch `p_burned` map.
* `11_CONSISTENCY_CHECKS/...` — consistency files, when enabled.

The Otsu-guided segmentation stage writes its outputs under the parallel
`DETERMINISTIC/<run_name>/` tree.
