---
title: "OtsuFire workflow overview"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{OtsuFire workflow overview}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
# eval = FALSE silences knitting. purl = FALSE is set per-chunk below
# instead of globally — opts_chunk$set(purl=FALSE) is not honoured by
# knitr::purl() because the setup chunk is extracted as text, not
# evaluated, during the R CMD check "running R code from vignettes"
# stage.
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>", eval = FALSE
)
```

This vignette describes the four-layer OtsuFire workflow at the API
level. Code chunks are shown but **not evaluated** because the
deterministic and supervised workflows operate on full-scale annual
mosaics (~Iberian Peninsula, 90 m resolution) that are too large to
ship with a CRAN package. The toy examples in
`vignette("OtsuFire", package = "OtsuFire")` cover the API surface
without external data.

## 1. Mosaic stage

`change_index_mosaic()` builds an annual change-index raster (RBR
or dNBR) from per-tile inputs. Output is a single GeoTIFF per year
that becomes the `change_index` argument to the deterministic /
supervised configs.

```{r mosaic, purl = FALSE}
library(OtsuFire)

mosaic_res <- change_index_mosaic(
  folder_path    = "/path/to/yearly/tiles",
  mask           = "/path/to/mask_Peninsula_3035.shp",
  year           = 2020L,
  raster_pattern = "MinMin_2020_tile*.tif",
  output_dir     = "/path/to/Composites_90m/Min_Min",
  nodata_value   = -9999,
  cap_below      = TRUE,
  lower_cap      = -1000,
  method         = "merge"
)

mosaic_path <- mosaic_res$mosaic_path
```

## 2. Deterministic stage

The deterministic workflow runs Otsu segmentation per CORINE x
ecoregion unit, grows seed-supported components, applies multi-stage
filtering, and emits an `internal_decisions.gpkg` layer.

```{r det, purl = FALSE}
det_cfg <- build_burned_mapping_config(
  change_index   = mosaic_path,
  vegetation_map = "/path/to/Corine_Masks/CLC_2018_peninsula.tif",
  burnable_mask  = "/path/to/Corine_Masks/burnable_mask_binary.tif",
  target_year    = 2020L,
  output_dir     = "/path/to/Results",
  run_name       = "rbr_2020_custom",
  detect_params = list(
    seed_threshold = 310,
    growth_delta = 90,
    minimum_growth_threshold = 240,
    minimum_seed_pixels = 30L
  ),
  refine_params = list(
    aoi_buffer_m = 5000,
    minimum_detected_area_m2 = 10,
    merge_overlaps = TRUE
  ),
  scoring_params = list(
    support_buffer_m = 0,
    previous_fire_exclusion_buffer_m = 90,
    previous_fire_cleanup_buffer_m = 90,
    minimum_remaining_area_m2 = 20000,
    reference_buffer_m = 90
  )
)

det_run <- run_deterministic_pipeline(det_cfg)
det_run$result_paths$internal_decisions
```

The deterministic builder no longer selects closed scenario presets.
Instead, users pass explicit ecological controls through
`detect_params`, `refine_params`, and `scoring_params`. Use the
`*_by_vegetation` keys when seed or growth behaviour must differ by
land-cover class; see `?build_burned_mapping_config` for the
vegetation-class mapping and the interpretation of each parameter.

If you need step-by-step access, the deterministic stage decomposes
into three modular helpers:

```{r det-modular, purl = FALSE}
patches  <- detect_burned_patches(det_cfg)
scored   <- score_burned_patches(patches, det_cfg)
# scored$internal_decisions is an sf with class_final, filter_1..3, etc.
```

## 3. Supervised one-year stage

The supervised stage trains an XGBoost model on the deterministic
decisions plus a burnable-only unburned negative pool, runs OOF
diagnostics, and emits a `final_map.gpkg` whose `p_burned` field is the
per-patch burned **model score**.

```{r sup, purl = FALSE}
sup_cfg <- build_supervised_burned_config(
  run_label          = "balanced",
  internal_decisions = det_run$result_paths$internal_decisions,
  change_index       = mosaic_path,
  hotspots           = "/path/to/Hotspots/hotspots_iberia_2020.geojson",
  target_year        = 2020L,
  output_dir         = "/path/to/Results",
  options = list(
    data_base      = "/path/to/Data",
    composite_base = "/path/to/Imagery/Composites_90m",
    result_name    = "Min_Min"
  )
)

sup_run <- run_oneyear_supervised_pipeline(
  config           = sup_cfg,
  run_consistency  = TRUE,
  overwrite        = FALSE
)
```

The supervised model uses **two explicit negative sources: random
background and Otsu residuals.** The negative-pool strategy is no longer
configurable. Every run uses `all_sources` implicitly: a burnable-only
unburned pool assembled from exactly two buckets — random burnable
background and Otsu current-year residual patches — each with its own
cap relative to the burned count. The two caps are the typed PUBLIC
block `negative_pool_params$caps = c(random = 1.0, otsu = 1.0)` (the
single source of truth; an `Inf` cap disables that bucket's cap),
mirrored into `cfg$train_control$caps`. The random-background RNG seed
lives in `cfg$train_control$seeds$random_seed` (builder argument
`random_seed`); the technical toggles live in `runtime_options =
list(reuse_existing = TRUE, write_outputs = TRUE, verbose = TRUE)` and do
not change which negatives are selected. The random background bucket
draws its sample (and computes its percentile) only within the burnable
domain, after the burnable mask and all exclusions/buffers; a location
that is both a random cell and an Otsu patch is kept once (Otsu has
priority). CORINE × ecoregion stratification is used only in the
deterministic delineation stage, not in the supervised negative pool.
The former user-selectable `negative_pool_policy` knob and its
`"deterministic_direct"` baseline were removed on 2026-06-05;
`negative_pool_policy = "all_sources"` is accepted as a no-op, and any
other value errors. The pipeline is entirely package-internal: no
external `sys.source()` of legacy scripts and no namespace mutation.

Training eligibility is defined by **explicit class, never by negation**.
A row with `class == "burned"` is a positive (always kept). A row with
`class == "unburned"` is a negative **only** when it resolves to exactly
one of the two valid buckets above (random, otsu); Otsu review / keep
patches are excluded and logged. Any other
case — `class` missing or unknown, or an unburned row with no valid
bucket — is a hard error. Review / keep / `NA` / unknown rows can never
silently become negatives (the old "negatives = everything that is not
burned" rule has been removed). The OOF folds and the FINAL model route
through ONE shared internal eligibility resolver and ONE shared capping
helper (`cap = ceiling(n_burned * ratio)`, take-all when available is at
or below the cap, all positives kept, negatives sampled without
replacement in a deterministic bucket order), so the two stages cannot
diverge on which rows are eligible or on how negatives are capped. The
only differences are the retained outer-test fold (OOF) and the seed.

The OOF diagnostics, the FINAL refit, and scoring share ONE feature
recipe (50 base features plus the matching `<feature>_isNA` indicators
created before imputation = 100 columns). A runtime feature-schema
parity guard aborts the run if the three feature spaces diverge.
OtsuFire uses a single training procedure: the number of boosting
rounds is selected with an inner validation split and early stopping,
then the recipe and model are refitted on all available training data
before prediction. There is no training-protocol choice.

`p_burned` (and the temporally adjusted `p_burned_current_year`) is a
model score under the training distribution — it must not be treated as
a calibrated probability without an explicit calibration step.

If you want explicit access to the supervised sub-stages,
`build_supervised_training_pools()`, `make_spatial_folds()`,
`extract_supervised_features()`, `run_oof_diagnostics()`,
`train_final_burned_model()`, and `score_supervised_burned_map()`
each cover one stage of the pipeline.

## 4. Validation and consistency

There are two distinct validations, against two different references.

**Internal validation (OOF).** The out-of-fold diagnostics
(`05_OOF`) measure how well the model reproduces the *deterministic
decision* labels, using spatial block folds. Under `nested_refit`, the
outer test fold is held out entirely (never used for the watchlist,
imputation fitting, selection, or the inner early-stopping split); the
model is fit on the outer-train rows, early-stopped on an inner
validation split, refit on all outer-train rows at the selected
`best_iteration`, and then predicts the untouched outer-test fold. OOF
AUC / PR-AUC are **not ground truth**: the labels are deterministic
decisions, not field observations.

**External validation.** `validate_fire_maps()` is
workflow-independent: threshold `p_burned` (or `p_burned_current_year`)
into a burned-area prediction layer and compare it against an external
reference (EFFIS). It reports omission, commission, F1, and IoU.

`validate_fire_maps()` compares an already-thresholded layer, so the
scored map is cut first. The operating threshold is a per-year decision
(`p_burned` is a score, not a calibrated probability); 0.5 below is only
a placeholder.

```{r validate, purl = FALSE}
fm <- sf::st_read(sup_run$final_map_gpkg, layer = "final_map_full", quiet = TRUE)
burned <- fm[is.finite(fm$p_burned_model) & fm$p_burned_model >= 0.5, ]

thresholded_gpkg <- "/path/to/Results/VALIDATION/thresholded_burned.gpkg"
sf::st_write(burned, thresholded_gpkg, layer = "thresholded_burned", quiet = TRUE)

val <- validate_fire_maps(
  input_shapefile = thresholded_gpkg,
  ref_shapefile   = "/path/to/Validation_fires_burneable_verano/Effis_CA_2020_maskKeep_summer.shp",
  mask_shapefile  = "/path/to/Mask_StudyArea/mask_Peninsula_3035.shp",
  burnable_raster = "/path/to/Corine_Masks/burnable_mask_binary.tif",
  year_target     = 2020L,
  validation_dir  = "/path/to/Results/VALIDATION",
  metrics_type    = "all"
)

val$metrics
```

External metrics are computed against a *different reference* (EFFIS)
than the OOF metrics (deterministic-decision labels). They are **not
comparable** to OOF AUC / PR-AUC — keep the two evaluations distinct.

**Temporal observability.** When an `observability_raster` (a day-of-year
raster) is supplied, observability is assessed **at the reference-fire
level**: a fire is either evaluated in full or excluded in full, and it is
never trimmed at pixel level. Two rules are available through
`observability_mode`.

`"legacy_any"` (the default, kept for reproducibility) excludes a fire whose
date falls after the last observation inside it, or that has no valid
observable DOY. Because the rule compares the fire date with the *maximum*
DOY in the perimeter, it is equivalent to requiring a single observable
pixel: a fire observed on 1 % of its area is kept whole. Excluded fires are
dropped from the reference *before* rasterization, so they contribute 0 TP /
0 FN, but their territory stays in the evaluation domain as reference = 0 —
a correct detection over such a fire is therefore scored as commission.

`"wholefire_fraction"` requires `temporal_observable_fraction >=
observability_min_fraction` (default `0.75`), the share of the fire's
burnable-domain pixels whose composite DOY is at or after the fire's
required DOY. The required DOY is taken from `ref_obs_doy_col` when that
argument is supplied and from `ref_end_doy_col` otherwise, with **no** silent
fallback to the start DOY: a fire with no usable date is reported as
`UNDETERMINED_OBS_DATE`.

`ref_obs_doy_col` is **authoritative**: when supplied it is the only date
column consulted, so an `NA` there means the reference is deliberately saying
"undetermined" and cannot be overridden by a stale `end_doy`. Source-specific
date semantics belong in the reference — the validator holds no per-source
rules. `ref_obs_doy_fallback = "end_doy"` opts back into a fallback and exists
only for reference layers whose observability column is populated for some
sources and not others. `undetermined_cause` records which situation produced
each `UNDETERMINED` fire, and a warning reports how many fires it affects.

**Evidence and domain membership are separate (2.3.0).** `observable_flag` is
the EVIDENCE flag and means only `observability_status == "OBSERVABLE"`; the
new `in_evaluation_domain` is the DOMAIN flag and is what drives the reference
subset and the masking. Missing authoritative observability metadata is
treated separately from evidence of temporal non-observability: under the
default `observability_undetermined_policy = "evaluate"`, fires with
`UNDETERMINED_OBS_DATE` remain positive reference features in the evaluation
domain and are flagged for subsequent audit. They contribute TP/FN like any
other reference fire and can never become background nor generate commission
through a missing date; no date is imputed. `"exclude"` removes them like a
non-observable fire and exists only for sensitivity runs.

Fires that DO carry evidence leave the domain entirely (reference, prediction
and strata set to `NA` over their geometry): `NOT_OBSERVABLE`, measured below
the threshold, and `NO_OBSERVABILITY_DATA`, with evaluable surface and a date
but no observation anywhere inside. `OUTSIDE_BURNABLE_DOMAIN` marks fires with
no burnable-domain cell at all — a domain fact, not a temporal failure.
Territory outside all of these is evaluated exactly as before, so false
positives away from reference fires are unaffected.
`observability_summary` and `OBSERVABILITY_SUMMARY_<year>.csv` report the
counts, the areas by status and the resulting `Evaluable_Area_ha`.

Two CSVs in `02_OBSERVABILITY/` carry the per-fire audit trail:
`REFERENCE_FIRES_DETECTION_<year>_<input>.csv` and, filtered to the
undetermined ones, `UNDETERMINED_FIRES_DETECTION_<year>_<input>.csv`. Each row
puts the observability verdict next to what the prediction actually did on that
fire (reference / detected / omitted cells and area, TP and FN, fire-level
recall), so an individual fire can be revisited later without recomputing the
validation.

Residual per-pixel non-observability inside kept fires is reported for audit
(`observability_audit`, `OBSERVABILITY_*` CSVs) but never changes any
metric. The reference cache is content-aware: it rebuilds automatically when
the observability raster, burnable mask, resolution, CRS, grid, study-area
mask, reference content or the observability rule — including the mode, the
threshold and `ref_obs_doy_col` — changes (`force_reprocess_ref = TRUE`
forces a rebuild).

### Corrected defect: OOF / FINAL `_isNA` divergence

A defect was found and corrected during Gate 1D.8, before the smoke run
and before the definitive run. Previously, **OOF generated the `_isNA`
missing-indicator companions, but the FINAL refit and scoring did
NOT** — so OOF evaluated a 102-column feature space while the deployed
model and scoring ran on a different (51-column) space. Consequently
the **prior OOF diagnostics did not exactly represent the deployed
model**, and the **prior FINAL was a different model** than the one OOF
evaluated (the prior FINAL was not necessarily invalid, just different).
The correction unified OOF, FINAL and scoring on the **same recipe**.
**Definitive results require retraining both models** under the
corrected recipe. Any comparison baseline produced under the current
code must be named the **"legacy training protocol under the corrected
common feature recipe"** — it is not an exact reproduction of the old
51-feature FINAL.

`check_supervised_consistency()` compares the deterministic and
supervised outputs of the same run and writes per-issue diagnostic
GPKG/CSV/TXT artefacts:

```{r consistency, purl = FALSE}
cons <- check_supervised_consistency(
  deterministic_decisions = det_run$internal_decisions_gpkg,
  final_map               = sup_run$final_map_gpkg,
  config                  = sup_cfg
)
```

The consistency check is also invoked automatically when
`run_oneyear_supervised_pipeline(..., run_consistency = TRUE)` is
called and `options$data_base` is supplied.

## Outputs

A single supervised one-year run for one (year, scenario) tuple
produces, under
`<output_dir>/<target_year>/<run_name>/SUPERVISED/<scenario>/`:

* `01_POOLS/...gpkg` — burned + unburned training pools.
* `02_FOLDS/...gpkg` — spatial folds for OOF.
* `03_FEATURES/features_geometry.gpkg` — full feature matrix.
* `04_MATRIX/...rds` — XGBoost design matrix snapshot.
* `05_OOF/..._oof_agg.csv` — out-of-fold diagnostics.
* `07_FINAL_MODEL_V2/..._final_model.rds` — trained model.
* `09_FINAL_MAP/..._final_map.gpkg` — final per-patch
  `p_burned` map.
* `11_CONSISTENCY_CHECKS/...` — consistency artefacts when enabled.

The deterministic stage writes its outputs under the parallel
`DETERMINISTIC/<run_name>/` tree.
