---
title: "OtsuFire workflow overview"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{OtsuFire workflow overview}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>", eval = FALSE
)
```

This vignette describes the four-layer OtsuFire workflow at the API
level. Code chunks are shown but **not evaluated** because the
deterministic and supervised workflows operate on full-scale annual
mosaics (~Iberian Peninsula, 90 m resolution) that are too large to
ship with a CRAN package. The toy examples in
`vignette("OtsuFire", package = "OtsuFire")` cover the API surface
without external data.

## 1. Mosaic stage

`change_index_mosaic()` builds an annual change-index raster (RBR
or dNBR) from per-tile inputs. Output is a single GeoTIFF per year
that becomes the `change_index` argument to the deterministic /
supervised configs.

```{r mosaic}
library(OtsuFire)

mosaic_path <- change_index_mosaic(
  input_dir   = "/path/to/yearly/tiles",
  output_dir  = "/path/to/Composites_90m/Min_Min",
  target_year = 2020L,
  index_type  = "RBR"
)
```

## 2. Deterministic stage

The deterministic workflow runs Otsu segmentation per CORINE x
ecoregion unit, grows seed-supported components, applies multi-stage
filtering, and emits an `internal_decisions.gpkg` layer.

```{r det}
det_cfg <- build_burned_mapping_config(
  scenario     = "balanced",
  change_index = mosaic_path,
  target_year  = 2020L,
  output_dir   = "/path/to/Results"
)

det_run <- run_deterministic_pipeline(det_cfg)
det_run$internal_decisions_gpkg
```

The `scenario` argument selects a per-CORINE-class parameter preset
(`balanced`, `original`, `lax`, `restrictive`). The presets are
loaded from `inst/extdata/` and are part of the validated
scientific defaults; they are **not** intended to be edited.

If you need step-by-step access, the deterministic stage decomposes
into three modular helpers:

```{r det-modular}
patches  <- detect_burned_patches(det_cfg)
scored   <- score_burned_patches(det_cfg, patches)
# scored is an sf with class_final, filter_1..3, etc.
```

## 3. Supervised one-year stage

The supervised stage trains an XGBoost model on the deterministic
decisions plus contextual unburned negatives, runs OOF diagnostics,
and emits a `final_map.gpkg` whose `p_burned` field is the per-patch
burned probability.

```{r sup}
sup_cfg <- build_supervised_burned_config(
  scenario           = "balanced",
  internal_decisions = det_run$internal_decisions_gpkg,
  change_index       = mosaic_path,
  hotspots           = "/path/to/Hotspots/hotspots_iberia_2020.geojson",
  target_year        = 2020L,
  output_dir         = "/path/to/Results",
  options = list(
    data_base            = "/path/to/Data",
    composite_base       = "/path/to/Imagery/Composites_90m",
    result_name          = "Min_Min",
    negative_pool_policy = "all_sources"
  )
)

sup_run <- run_oneyear_supervised_pipeline(
  config           = sup_cfg,
  run_consistency  = TRUE,
  overwrite        = FALSE
)
```

`negative_pool_policy` controls how unburned negatives are sourced:

* `"deterministic_direct"` — Mode A baseline. Uses only the
  deterministic decisions to derive contextual exclusions, spectral
  hard negatives, random burnable background, and Otsu-unburned
  samples.
* `"all_sources"` — operational default for the `balanced`,
  `restrictive`, and `strict` scenarios. Adds a current-year Otsu
  unburned pool computed by the package-internal legacy helpers
  (`polygonize_Otsu()`, `coverage_by_patch_raster()`,
  `run_scenarios()`, `process_otsu_rasters_()`).

Both modes are entirely package-internal: no external `sys.source()`
of legacy scripts and no namespace mutation are performed.

If you want explicit access to the supervised sub-stages,
`build_supervised_training_pools()`, `make_spatial_folds()`,
`extract_supervised_features()`, `run_oof_diagnostics()`,
`train_final_burned_model()`, `score_supervised_burned_map()`, and
`update_burned_like_registry()` each cover one stage of the
pipeline.

## 4. Validation and consistency

`validate_fire_maps()` is workflow-independent: it can be applied to
any compatible burned-area prediction layer, including supervised
outputs after thresholding `p_burned`.

```{r validate}
val <- validate_fire_maps(
  predicted = sup_run$final_map_gpkg,
  reference = "/path/to/Validation_fires_burneable_verano/Effis_CA_2020_maskKeep_summer.shp",
  threshold = 0.5
)
```

`check_supervised_consistency()` compares the deterministic and
supervised outputs of the same run and writes per-issue diagnostic
GPKG/CSV/TXT artefacts:

```{r consistency}
cons <- check_supervised_consistency(
  deterministic_decisions = det_run$internal_decisions_gpkg,
  final_map               = sup_run$final_map_gpkg,
  config                  = sup_cfg
)
```

The consistency check is also invoked automatically when
`run_oneyear_supervised_pipeline(..., run_consistency = TRUE)` is
called and `options$data_base` is supplied.

## Outputs

A single supervised one-year run for one (year, scenario) tuple
produces, under
`<output_dir>/<target_year>/<run_name>/SUPERVISED/<scenario>/`:

* `01_POOLS/...gpkg` — burned + unburned training pools.
* `02_FOLDS/...gpkg` — spatial folds for OOF.
* `03_FEATURES/features_geometry.gpkg` — full feature matrix.
* `04_MATRIX/...rds` — XGBoost design matrix snapshot.
* `05_OOF/..._oof_agg.csv` — out-of-fold diagnostics.
* `07_FINAL_MODEL_V2/..._final_model.rds` — trained model.
* `09_FINAL_MAP/..._final_map.gpkg` — final per-patch
  `p_burned` map.
* `11_CONSISTENCY_CHECKS/...` — consistency artefacts when enabled.

The deterministic stage writes its outputs under the parallel
`DETERMINISTIC/<scenario>/` tree.
