---
title: "OtsuFire workflow overview"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{OtsuFire workflow overview}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
# eval = FALSE silences knitting. purl = FALSE is set per-chunk below
# instead of globally — opts_chunk$set(purl=FALSE) is not honoured by
# knitr::purl() because the setup chunk is extracted as text, not
# evaluated, during the R CMD check "running R code from vignettes"
# stage.
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>", eval = FALSE
)
```

This vignette describes the four-layer OtsuFire workflow at the API
level. Code chunks are shown but **not evaluated** because the
deterministic and supervised workflows operate on full-scale annual
mosaics (~Iberian Peninsula, 90 m resolution) that are too large to
ship with a CRAN package. The toy examples in
`vignette("OtsuFire", package = "OtsuFire")` cover the API surface
without external data.

## 1. Mosaic stage

`change_index_mosaic()` builds an annual change-index raster (RBR
or dNBR) from per-tile inputs. Output is a single GeoTIFF per year
that becomes the `change_index` argument to the deterministic /
supervised configs.

```{r mosaic, purl = FALSE}
library(OtsuFire)

mosaic_path <- change_index_mosaic(
  input_dir   = "/path/to/yearly/tiles",
  output_dir  = "/path/to/Composites_90m/Min_Min",
  target_year = 2020L,
  index_type  = "RBR"
)
```

## 2. Deterministic stage

The deterministic workflow runs Otsu segmentation per CORINE x
ecoregion unit, grows seed-supported components, applies multi-stage
filtering, and emits an `internal_decisions.gpkg` layer.

```{r det, purl = FALSE}
det_cfg <- build_burned_mapping_config(
  change_index   = mosaic_path,
  vegetation_map = "/path/to/Corine_Masks/CLC_2018_peninsula.tif",
  burnable_mask  = "/path/to/Corine_Masks/burnable_mask_binary.tif",
  target_year    = 2020L,
  output_dir     = "/path/to/Results",
  run_name       = "rbr_2020_custom",
  detect_params = list(
    seed_threshold = 310,
    growth_delta = 90,
    minimum_growth_threshold = 240,
    minimum_seed_pixels = 30L
  ),
  refine_params = list(
    aoi_buffer_m = 5000,
    minimum_detected_area_m2 = 10,
    merge_overlaps = TRUE
  ),
  scoring_params = list(
    support_buffer_m = 0,
    previous_fire_exclusion_buffer_m = 90,
    previous_fire_cleanup_buffer_m = 90,
    minimum_remaining_area_m2 = 20000,
    reference_buffer_m = 90
  )
)

det_run <- run_deterministic_pipeline(det_cfg)
det_run$internal_decisions_gpkg
```

The deterministic builder no longer selects closed scenario presets.
Instead, users pass explicit ecological controls through
`detect_params`, `refine_params`, and `scoring_params`. Use the
`*_by_vegetation` keys when seed or growth behaviour must differ by
land-cover class; see `?build_burned_mapping_config` for the
vegetation-class mapping and the interpretation of each parameter.

If you need step-by-step access, the deterministic stage decomposes
into three modular helpers:

```{r det-modular, purl = FALSE}
patches  <- detect_burned_patches(det_cfg)
scored   <- score_burned_patches(det_cfg, patches)
# scored is an sf with class_final, filter_1..3, etc.
```

## 3. Supervised one-year stage

The supervised stage trains an XGBoost model on the deterministic
decisions plus a burnable-only unburned negative pool, runs OOF
diagnostics, and emits a `final_map.gpkg` whose `p_burned` field is the
per-patch burned **model score**.

```{r sup, purl = FALSE}
sup_cfg <- build_supervised_burned_config(
  scenario           = "balanced",
  internal_decisions = det_run$internal_decisions_gpkg,
  change_index       = mosaic_path,
  hotspots           = "/path/to/Hotspots/hotspots_iberia_2020.geojson",
  target_year        = 2020L,
  output_dir         = "/path/to/Results",
  options = list(
    data_base      = "/path/to/Data",
    composite_base = "/path/to/Imagery/Composites_90m",
    result_name    = "Min_Min"
  )
)

sup_run <- run_oneyear_supervised_pipeline(
  config           = sup_cfg,
  run_consistency  = TRUE,
  overwrite        = FALSE
)
```

The negative-pool strategy is no longer configurable. Every run uses
`all_sources` implicitly: a burnable-only, four-bucket unburned
taxonomy (contextual exclusions, spectral hard negatives, random
burnable background, and Otsu current-year residual patches) with
independent caps relative to the burned count. The random background
bucket draws its sample (and computes its percentile) only within the
burnable domain, after the burnable mask and all exclusions/buffers.
CORINE × ecoregion stratification is used only in the deterministic
delineation stage, not in the supervised negative pool. The former
user-selectable `negative_pool_policy` knob and its
`"deterministic_direct"` baseline were removed on 2026-06-05;
`negative_pool_policy = "all_sources"` is accepted as a no-op, and any
other value errors. The pipeline is entirely package-internal: no
external `sys.source()` of legacy scripts and no namespace mutation.

Training eligibility is defined by **explicit class, never by negation**.
A row with `class == "burned"` is a positive (always kept). A row with
`class == "unburned"` is a negative **only** when it resolves to exactly
one of the four valid buckets above (contextual, spectral, random,
otsu); Otsu review / keep patches are excluded and logged. Any other
case — `class` missing or unknown, or an unburned row with no valid
bucket — is a hard error. Review / keep / `NA` / unknown rows can never
silently become negatives (the old "negatives = everything that is not
burned" rule has been removed). The OOF folds and the FINAL model route
through ONE shared internal eligibility resolver and ONE shared capping
helper (`cap = ceiling(n_burned * ratio)`, take-all when available is at
or below the cap, all positives kept, negatives sampled without
replacement in a deterministic bucket order), so the two stages cannot
diverge on which rows are eligible or on how negatives are capped. The
only differences are the retained outer-test fold (OOF) and the seed.

The OOF diagnostics, the FINAL refit, and scoring share ONE feature
recipe (50 base features plus the matching `<feature>_isNA` indicators
created before imputation = 100 columns). A runtime feature-schema
parity guard aborts the run if the three feature spaces diverge.
OtsuFire uses a single training procedure: the number of boosting
rounds is selected with an inner validation split and early stopping,
then the recipe and model are refitted on all available training data
before prediction. There is no training-protocol choice.

`p_burned` (and the temporally adjusted `p_burned_current_year`) is a
model score under the training distribution — it must not be treated as
a calibrated probability without an explicit calibration step.

If you want explicit access to the supervised sub-stages,
`build_supervised_training_pools()`, `make_spatial_folds()`,
`extract_supervised_features()`, `run_oof_diagnostics()`,
`train_final_burned_model()`, and `score_supervised_burned_map()`
each cover one stage of the pipeline.

## 4. Validation and consistency

There are two distinct validations, against two different references.

**Internal validation (OOF).** The out-of-fold diagnostics
(`05_OOF`) measure how well the model reproduces the *deterministic
decision* labels, using spatial block folds. Under `nested_refit`, the
outer test fold is held out entirely (never used for the watchlist,
imputation fitting, selection, or the inner early-stopping split); the
model is fit on the outer-train rows, early-stopped on an inner
validation split, refit on all outer-train rows at the selected
`best_iteration`, and then predicts the untouched outer-test fold. OOF
AUC / PR-AUC are **not ground truth**: the labels are deterministic
decisions, not field observations.

**External validation.** `validate_fire_maps()` is
workflow-independent: threshold `p_burned` (or `p_burned_current_year`)
into a burned-area prediction layer and compare it against an external
reference (EFFIS). It reports omission, commission, F1, and IoU.

```{r validate, purl = FALSE}
val <- validate_fire_maps(
  predicted = sup_run$final_map_gpkg,
  reference = "/path/to/Validation_fires_burneable_verano/Effis_CA_2020_maskKeep_summer.shp",
  threshold = 0.5
)
```

External metrics are computed against a *different reference* (EFFIS)
than the OOF metrics (deterministic-decision labels). They are **not
comparable** to OOF AUC / PR-AUC — keep the two evaluations distinct.

### Corrected defect: OOF / FINAL `_isNA` divergence

A defect was found and corrected during Gate 1D.8, before the smoke run
and before the definitive run. Previously, **OOF generated the `_isNA`
missing-indicator companions, but the FINAL refit and scoring did
NOT** — so OOF evaluated a 102-column feature space while the deployed
model and scoring ran on a different (51-column) space. Consequently
the **prior OOF diagnostics did not exactly represent the deployed
model**, and the **prior FINAL was a different model** than the one OOF
evaluated (the prior FINAL was not necessarily invalid, just different).
The correction unified OOF, FINAL and scoring on the **same recipe**.
**Definitive results require retraining both models** under the
corrected recipe. Any comparison baseline produced under the current
code must be named the **"legacy training protocol under the corrected
common feature recipe"** — it is not an exact reproduction of the old
51-feature FINAL.

`check_supervised_consistency()` compares the deterministic and
supervised outputs of the same run and writes per-issue diagnostic
GPKG/CSV/TXT artefacts:

```{r consistency, purl = FALSE}
cons <- check_supervised_consistency(
  deterministic_decisions = det_run$internal_decisions_gpkg,
  final_map               = sup_run$final_map_gpkg,
  config                  = sup_cfg
)
```

The consistency check is also invoked automatically when
`run_oneyear_supervised_pipeline(..., run_consistency = TRUE)` is
called and `options$data_base` is supplied.

## Outputs

A single supervised one-year run for one (year, scenario) tuple
produces, under
`<output_dir>/<target_year>/<run_name>/SUPERVISED/<scenario>/`:

* `01_POOLS/...gpkg` — burned + unburned training pools.
* `02_FOLDS/...gpkg` — spatial folds for OOF.
* `03_FEATURES/features_geometry.gpkg` — full feature matrix.
* `04_MATRIX/...rds` — XGBoost design matrix snapshot.
* `05_OOF/..._oof_agg.csv` — out-of-fold diagnostics.
* `07_FINAL_MODEL_V2/..._final_model.rds` — trained model.
* `09_FINAL_MAP/..._final_map.gpkg` — final per-patch
  `p_burned` map.
* `11_CONSISTENCY_CHECKS/...` — consistency artefacts when enabled.

The deterministic stage writes its outputs under the parallel
`DETERMINISTIC/<run_name>/` tree.
